Causal Inference with Graphical Neural Networks

### 2024-04-16 latex/draft.tex
Pushing latex changes.

\documentclass{article}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{ntheorem}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}
\usepackage{tkz-graph}
\usetikzlibrary{arrows}
\usetikzlibrary{shapes.geometric}
\usepackage{hyperref}

@@

\begin{align*}
    P(\text{LC} = \text{T} \;&|\; \textbf{do}(\text{YF} = \text{T})), \\
    P(\text{LC} = \text{T} \;&|\; \textbf{do}(\text{PP} = \text{T})), \\
    P(\text{LC} = \text{T} \;&|\; \textbf{do}(\text{A} = \text{T})), \\
    P(\text{LC} = \text{T} \;&|\; \textbf{do}(\text{AD} = \text{T})), \\
    P(\text{LC} = \text{T} \;&|\; \textbf{do}(\text{CA} = \text{T})), \\
\end{align*}

@@

Plugging forward with \textbf{do} calculus, we can write $P(\text{LC} = \text{T} \;|\; \textbf{do}(\text{YF} = \text{T}))$ as
(using corollary~\ref{corollary:do})

\begin{equation*}
    P(\text{LC} = \text{T} \;|\; \textbf{do}(\text{YF} = \text{T})) = 
    \sum_{s\in\{T, F\}} \frac{P(\text{YF} = \text{T},\, \text{LC} = \text{T},\, \text{S} = s)}{P(\text{YF} = \text{T}\,|\, \text{S} = s)},
\end{equation*}

where $\text{S}$ is smoking. Using the property that we can split the numerator into probabilities only involving parents ({\color{red}rewrite})

\begin{equation*}
    \sum_{s\in\{T, F\}}\frac{P(\text{YF} =\text{T}\;|\; \text{S} = s) P(\text{LC} 
    = \text{T}\;|\; \text{S}=s,\, \text{G}) P(\text{S}=s\;|\;\text{A},\, \text{PP})}
    {P(\text{YF} = \text{T}\;|\; \text{S} = s)}.
\end{equation*}

All of these fun conditionals lead us to

\begin{equation*}
    \sum_{s,a,g,p}\frac{P(\text{YF} =\text{T}\;|\; \text{S} = s) P(\text{LC} 
    = \text{T}\;|\; \text{S}=s,\, \text{G}=g) P(\text{S}=s\;|\;\text{A}=a,\, \text{PP}=p)}
    {P(\text{YF} = \text{T}\;|\; \text{S} = s)},
\end{equation*}

where each $s,a,g,p$ is summed over the set $\{T, F\}$.

@@

>>>>

\documentclass{article}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{ntheorem}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}
\usepackage{tkz-graph}
\usetikzlibrary{arrows}
\usetikzlibrary{shapes.geometric}
\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}
\usepackage{hyperref}

@@

\begin{align*}
    P(\text{LC} = \text{T} \;&|\; \textbf{do}(\text{YF} = \text{T})), \\
    P(\text{LC} = \text{T} \;&|\; \textbf{do}(\text{PP} = \text{T})), \\
    P(\text{LC} = \text{T} \;&|\; \textbf{do}(\text{A} = \text{T})), \numberthis \label{do_list}\\
    P(\text{LC} = \text{T} \;&|\; \textbf{do}(\text{AD} = \text{T})), \\
    P(\text{LC} = \text{T} \;&|\; \textbf{do}(\text{CA} = \text{T})), \\
\end{align*}

@@

Plugging forward with \textbf{do} calculus, we can do one as an example. We write $P(\text{LC} = \text{T} \;|\; \textbf{do}(\text{YF} = \text{T}))$
(using corollary~\ref{corollary:do}) as

\begin{equation}
    P(\text{LC} = \text{T} \;|\; \textbf{do}(\text{YF} = \text{T})) = 
    \sum_{s\in\{T, F\}} \frac{P(\text{YF} = \text{T},\, \text{LC} = \text{T},\, \text{S} = s)}{P(\text{YF} = \text{T}\,|\, \text{S} = s)},
    \label{yf1}
\end{equation}

where $\text{S}$ is smoking. Since the Lung Cancer node has Smoking and Genetics as parents, and Smoking
has parent Anxiety and Peer Pressure (see Figure~\ref{fig:lucas_dag}), we can write Equation~\ref{yf1} to

\begin{equation}
    \sum_{S, G, A, PP}\frac{P(\text{YF} =\text{T}\;|\; \text{S}) P(\text{LC} 
    = \text{T}\;|\; \text{S},\, \text{G}) P(\text{G}) P(\text{S}\;|\;\text{A},\, \text{PP}) P(\text{A})P(\text{PP})}
    {P(\text{YF} = \text{T}\;|\; \text{S})}.
    \label{yf2}
\end{equation}

where each $\text{S},\text{A}, \text{G}, \text{PP}$ is in terms of $\{T, F\}$.

Since not every term is in terms of every variable, we can re-write the Equation~\ref{yf2} as

\begin{align*}
    &=\sum_{S} \frac{P(\text{YF} = T\,|\, S)}{P(\text{YF} = T\,|\, \text{S})}
        \Biggl(\sum_{G} P(\text{LC} = \text{T}\,|\, \text{S}, \text{G})P(\text{G})\Biggr)
            \Biggl(\sum_{A} P(\text{A}) 
                \Bigl(\sum_{\text{PP}}P(\text{S}\,|\, \text{A}, \text{PP})P(\text{PP})\Bigr) 
            \Biggr) \\
    &=\sum_{S}\Big(\sum_{G} P(\text{LC} = \text{T}\,|\, \text{S}, \text{G})P(\text{G})\Big)
        \Biggl(\sum_{A} P(\text{A}) 
            \Bigl(\sum_{\text{PP}}P(\text{S}\,|\, \text{A}, \text{PP})P(\text{PP})\Bigr) 
        \Biggr). \numberthis \label{yf3}
\end{align*}

Using the probabilities from Table~\ref{tab:lucas_probs} on Equation~\ref{yf3} we get

\begin{equation*}
    P(\text{LC} = \text{T} \;|\; \textbf{do}(\text{YF} = \text{T})) = {\color{red} \rule{2cm}{0.15mm} \text{ fix this!!}}
\end{equation*}

Performing similar calculations on the other probabilities in Equation~\ref{do_list}, we get

\begin{table}[h]
    \centering
    \begin{tabular}{|c|c|}
        \hline
        \textit{do Statements} & \textit{Probabilities} \\ \hline
        $P$(\text{LC} = \text{T} $\vert$ \textbf{do}(\text{YF} = \text{T})) &  \\ \hline
        $P$(\text{LC} = \text{T} $\vert$ \textbf{do}(\text{PP} = \text{T})) &  \\ \hline
        $P$(\text{LC} = \text{T} $\vert$ \textbf{do}(\text{A} = \text{T})) &   \\ \hline
        $P$(\text{LC} = \text{T} $\vert$ \textbf{do}(\text{AD} = \text{T})) &  \\ \hline
        $P$(\text{LC} = \text{T} $\vert$ \textbf{do}(\text{CA} = \text{T})) &  \\ \hline
    \end{tabular}
    \caption{The resulting probabilities of the \textbf{do} statements found in Equation~\ref{do_list}.
    For a more detailed workthrough of these probabilities, see Appendix~\ref{do_work}.}
    \label{tab:do_work_tab}
\end{table}

% \begin{equation}

% \end{equation}

@@

\section{Conditioning Outside Markov Blanket Results}
\label{do_work}

\subsection{$P$(\text{LC} = \text{T} \,$\vert$\, \textbf{do}(\text{PP} = \text{T}))}

We have $P(\text{LC} = \text{T} \,\vert\, \textbf{do}(\text{PP} = \text{T}))$ yields

\begin{align*}
     &= P(\text{LC} = \text{T},\, \text{PP} = \text{T}) \\
        &= P(\text{PP} = \text{T})\left(\sum_{s, g, an}P(\text{LC} = T \,|\, \text{S} = s,\, \text{G} = g)P(\text{G} = g)P(\text{An} = an)\right) \\
        &= P(\text{PP} = \text{T})\Big(\sum_{an}P(\text{An} = an)\Big)\left(\sum_{s, g}P(\text{LC} = T \,|\, \text{S} = s,\, \text{G} = g)P(\text{G} = g)\right) \\
        &= P(\text{PP} = \text{T})\left(\sum_{s, g}P(\text{LC} = T \,|\, \text{S} = s,\, \text{G} = g)P(\text{G} = g)\right) \\
    &= \rule{2cm}{0.15mm}.
\end{align*}

\subsection{$P$(\text{LC} = \text{T} $\vert$ \textbf{do}(\text{An} = \text{T}))}

We have $P(\text{LC} = \text{T} \,\vert\, \textbf{do}(\text{An} = \text{T}))$ yields

\begin{align*}
     &= P(\text{LC} = \text{T},\, \text{An} = \text{T}) \\
        &= P(\text{An} = \text{T})\left(\sum_{s, g, p}P(\text{LC} = T \,|\, \text{S} = s,\, \text{G} = g)P(\text{G} = g)P(\text{S}=s\,|\, \text{An} = T, \text{PP} = p)P(\text{PP} = p)\right) \\
    &= \rule{2cm}{0.15mm}.
\end{align*}

\subsection{$P$(\text{LC} = \text{T} $\vert$ \textbf{do}(\text{AD} = \text{T}))}

We have

\begin{align*}
    P(\text{LC} = \text{T} \,\vert\, \textbf{do}(\text{AD} = \text{T})) &= P(\text{LC} = \text{T},\, \text{AD} = \text{T},\, \text{G}) \\
        &= \Biggl( \sum_{g} P(\text{AD} = \text{T}\,\vert\, \text{G} = g)\Bigl(\sum_{s}P(\text{LC} = T \,|\, \text{S} = s,\, \text{G} = g)\Bigr)\Biggr) \\
    &= \rule{2cm}{0.15mm}.
\end{align*}
